spinifex: manual control of dynamic linear projections of high-dimensional data

### 2019-04-11 paper/spinifex-paper.tex
Knit of last changes

\abstract{%
The set of dynamic linear projections of multivariate data collectively
known as \textbf{tours} provides an important tool for extending the
dimensionality of visuals. The R package \textbf{tourr} offers a variety
of path generators and geometric displays for conducting tours. This
paper discusses an extension package, \textbf{spinifex}, that adds
support for the path generation of manual tours and extends the display
of tours to use with the animation packages, \textbf{plotly} and
\textbf{gganimate}. Manual tours are used to explore the sensitivity of
structure as the contributions of a manipulation variable are changed. A
recent paper (Wang et al.
\protect\hyperlink{ref-wang_mapping_2018}{2018}) visualizes the
sensitivity of the hadronic experiments to nucleon structure.
Sensitivity was characterized in non-linear 3D embeddings of the first
10 principal components. This research applies manual tours to this data
showing that manual tours resolve information about cluster structure
and identify outliers, information that is orthogonal to the original
static viewing plane.
}

@@

A tour is a multivariate data analysis technique in which a sequence of
linear (orthogonal) projections are viewed as an animation while the
orientation of the projection basis is rotated across time. Each frame
of the sequence corresponds to a small change in the projection for a
smooth transition that perseveres continuity. This is a crucial tool
extending the dimensionality of visualization which is a large boon as
datasets and models exist in increasing cardinality. The package
\CRANpkg{tourr} (Wickham et al.
\protect\hyperlink{ref-wickham_tourr_2011}{2011}) builds a platform for
generating tour paths and applying to a wide number of display
geometrics in \textbf{base} graphics.

While there are numerous methods that generate tour paths, this research
focuses on the manual tour. The manual tour was described in Cook and
Buja (\protect\hyperlink{ref-cook_manual_1997}{1997}) and allows a user
to control the projection coefficients of a selected variable in a 2D
projection. The manipulation of these coefficients allows the analyst to
explore how sensitive the projections structure is to these changes.
This makes manual tours particularly useful once a feature of interest
has been identified, for example, with the use of a guided tour (Hurley
and Buja \protect\hyperlink{ref-hurley_analyzing_1990}{1990}). Rather
than exploring the large phase space of the data, a guided tour selects
a very specific path via projection pursuit (Diaconis and Freedman
\protect\hyperlink{ref-diaconis_asymptotics_1984}{1984}), the
optimization of an index function on the projection via a hill climbing
algorithm. This allows guided tours to rapidly identify interesting
projection features given the relatively large parameter-space. After a
projection of interest is identified an analyst can then use the ``finer
brush'' of the manual tour, by controlling the contributions of
individual variables.

\CRANpkg{spinifex} utilizes two new animation packages, \CRANpkg{plotly}
(Sievert \protect\hyperlink{ref-sievert_plotly_2018}{2018}) and
\CRANpkg{gganimate} (Pedersen and Robinson
\protect\hyperlink{ref-pedersen_gganimate:_2019}{2019}), to display
tours, manual or other saved tours. From a given projection, the user
can choose which variable to control, and the animation sequence is
generated to remove the variable from the projection, and then extend
its contribution to be the sole variable in one direction. This allows
the viewer to assess the change in structure induced in the projection
by the variable's contribution.

The paper is organized as follows. Section \ref{sec:algorithm} explains
the algorithm using a toy dataset. Section \ref{sec:display} discussed
the display of the animation after the path has been generated. Section
\ref{sec:application} illustrates how this can be used for sensitivity
analysis applied to contemporary high energy physics. The last section,
\ref{sec:discussion} summarizes the work and discusses future research.

@@

The flea data, originally from Lubischew
(\protect\hyperlink{ref-lubischew_use_1962}{1962}), made available in
\textbf{tourr} is used to illustrate the algorithm. The data contains 74
observations across 6 variables, physical measurements of the flea
beetles. Each observation belonging to one of three species.

The data is defined. A basis set (ideally orienting to an interesting
feature) must be provided as an initial orientation. One way to identify
a projection containing interesting features is to apply a guided tour
(Cook, Swayne, and Buja
\protect\hyperlink{ref-cook_interactive_2007}{2007}). In a guided tour
the projection sequence is selected by optimizing an index function via
hill-climbing on the projection space. In this case, the holes index is
selected and applied to standardized flea data. The holes index is
maximized when the projected observations are furthest from the center.
Figure \ref{fig:step0} shows a locally optimized projection of the data.
The left panel displays the reference axes of the projection basis, a
visual indication of the magnitude and direction each variable
contributed to the projection. The right panel shows the data as
projected through the basis set described by the reference axes (left).
Data points are colored and given shape according to the species (while
the guided tour was unsupervised with this information).

@@

\caption[Basis reference axes (left) and projected data (right) of standardized flea data]{Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. Basis identified by a holes-index guided tour. The variables aede2 and tars1 contribute mostly orthogonal to the other variables.}\label{fig:step0}
\end{figure}
\end{Schunk}

@@

(ref:JetClusterBad-cap) A radial manual tour of PC4 within the jet
cluster. Colored by experiment type: ATLAS7new in green and ATLAS7old in
orange. Data from ATLAS7new (green) spans mostly the same space as
ALTLAS7old (orange) with no evident difference in cluster structure
across varying contributions of PC3. A dynamic version can be viewed at
\url{https://nspyrison.netlify.com/thesis/jetcluster_manualtour_pc3/}.

@@

\caption[(ref:JetClusterBad-cap)]{(ref:JetClusterBad-cap)}\label{fig:JetClusterBad}
\end{figure}
\end{Schunk}

@@

A different space is used to explore the DIS cluster; specifically the
first six principal components, which explain 48\% of the variation
contained within the aggregated data(Cook, Laa, and Valencia
\protect\hyperlink{ref-cook_dynamical_2018}{2018}). Radial manual tours
are performed on PC6 and PC2 in figure \ref{fig:DISclusterGood} and
figure \ref{fig:DISclusterBad} respectively.

The selection of the manip variable is important as the manipulation
spaces convey substantially different information. The manual tour of
PC6 offers information about the dimensionality, shape, and orientations
of the different experiment classes. whereas manipulating the
contributions of PC2 only shows a subset of the dimensionality and shape
information. Manipulating the contributions of PC6 turned out to be much
more insightful than that of PC2. This result might seem
counter-intuitive at first as PC2 should explain much more of the
variation in the data. However, features and structure in the data
regularly reside in finer detail that can be lost when looking only at
static projections.

@@

Tours, the dynamic linear projection of multivariate data, play an
important role in data visualization; they extending the dimensionality
of visuals while data- and parameter-spaces become ever larger. This
research has modified the algorithm producing manual tours and has made
this functionality available in package \textbf{spinifex}. The package
adds to \textbf{tourr}, extending the graphics offerings that can be
used to display tours.

Radial manual tours were applied to a dataset across different
experiments of hadronic collisions. The importance of selecting the
correct variable to manipulate is demonstrated by comparing tours of
varying quality. The manual tours convey a better picture of the
structure of the clustering than static linear or non-linear projections
are able to. Giving the full contribution of the manipulation space to
the manip var enables analysts to explore the sensitivity of the
structure with the selected variable. This information can be used by
physicists to which experiments are probing which parameter spaces,
which can be indispensable for smarter planning of effort and funding.

@@

The \textbf{tourr} package provides many other geometric displays with
the \texttt{tourr::display\_*()} family. These geometric options could
be integrated into the \textbf{ggplot2} framework for display on
\textbf{plotly} and \textbf{gganimate}. Additionally, the
\CRANpkg{animation} package Xie et al.
(\protect\hyperlink{ref-xie_animation:_2018}{2018}) could be implemented
for another graphics framework. However, \textbf{animation} builds from
\textbf{base} graphs while \textbf{spinifex} utilizes \textbf{ggplot2}
graphics.

@@

\leavevmode\hypertarget{ref-cook_interactive_2007}{}%
Cook, Dianne, Deborah F. Swayne, and A. Buja. 2007. \emph{Interactive
and Dynamic Graphics for Data Analysis: With R and GGobi}. Springer
Science \& Business Media.

\leavevmode\hypertarget{ref-diaconis_asymptotics_1984}{}%
Diaconis, Persi, and David Freedman. 1984. ``Asymptotics of Graphical
Projection Pursuit.'' \emph{The Annals of Statistics} 12 (3): 793--815.

@@

>>>>

\abstract{%
The set of dynamic linear projections of multivariate data collectively
known as \textbf{tours} provides an important tool for extending the
dimensionality of visuals. The R package \textbf{tourr} offers a variety
of path generators and geometric displays for conducting tours. This
paper discusses an extension package, \textbf{spinifex}, that adds
support for the path generation of manual tours and extends the display
of tours to use with the animation packages, \textbf{plotly} and
\textbf{gganimate}. Manual tours are used to explore the sensitivity of
structure as the contributions of a manipulation variable are changed. A
recent paper (Wang et al.
\protect\hyperlink{ref-wang_mapping_2018}{2018}) visualized the
sensitivity of the hadronic experiments to nucleon structure.
Sensitivity was characterized in non-linear 3D embeddings of the first
10 principal components. This research applies manual tours to this data
showing that manual tours resolve information about the structure of
cluster distribution by bringing to light information that is orthogonal
to static viewing planes.
}

@@

A tour is a multivariate data analysis technique in which a sequence of
linear (orthogonal) projections are viewed as an animation while the
orientation of the projection basis is rotated across time. Each frame
of the sequence corresponds to a small change in the projection for a
smooth transition that preserves continuity. Tours are a crucial tool
extending the dimensionality of visualization which is increasingly
important as datasets and models exist in increasing cardinality. The
package \CRANpkg{tourr} (Wickham et al.
\protect\hyperlink{ref-wickham_tourr_2011}{2011}) builds a platform for
generating tour paths and applying to a wide number of display
geometrics in \textbf{base} graphics.

While there are numerous methods that generate tour paths, this research
focuses on the manual tour. The manual tour was described in Cook and
Buja (\protect\hyperlink{ref-cook_manual_1997}{1997}) and allows a user
to control the projection coefficients of a selected variable in a 2D
projection. The manipulation of these coefficients allows the analyst to
explore how sensitive the projection structure is to these
manipulations. As manual tours operate on only one variable at a time,
they particularly useful once a feature of interest has been identified.
One way to identify ``interesting'' features is with the use of a guided
tour (Hurley and Buja
\protect\hyperlink{ref-hurley_analyzing_1990}{1990}). Guided tours
select a very specific path, that which approachs a projection that
optimizes an objective-function, similar to simulated annealing
(Kirkpatrick, Gelatt, and Vecchi
\protect\hyperlink{ref-kirkpatrick_optimization_1983}{1983}). The direct
optimization of a function allows guided tours to rapidly identify
interesting projection features given the relatively large
parameter-space. After a projection of interest is identified an analyst
can then use the ``finer brush'' of the manual tour, by controlling the
contributions of individual variables to explore the sensitivity they
have to the projection structure.

This paper offers an implementation of manual tours with the package
\CRANpkg{spinifex}. This package also exends the graphics display of
\textbf{tourr} to the animation frameworks offered by \CRANpkg{plotly}
(Sievert \protect\hyperlink{ref-sievert_plotly_2018}{2018}) and
\CRANpkg{gganimate} (Pedersen and Robinson
\protect\hyperlink{ref-pedersen_gganimate:_2019}{2019}). The algorithm
used to perform a radial manual tour is demonstrated on a toy data set.
Followed by an application and disscussion of the radial manual tour on
an aggregation of high energy physics experiments in a large
multivariate space.

The paper is organized as follows. Section \ref{sec:algorithm} explains
the algorithm using a toy dataset. Section \ref{sec:display} discusses
the display of the animation after the path has been generated. Section
\ref{sec:application} illustrates how this can be used for sensitivity
analysis applied to contemporary high energy physics. The last section,
\ref{sec:discussion} summarizes the work and discusses future research.

@@

The flea data, originally from Lubischew
(\protect\hyperlink{ref-lubischew_use_1962}{1962}), made available in
\textbf{tourr} is used to illustrate the algorithm. The data contains 74
observations across 6 variables, which are physical measurements of the
flea beetles. Each observation belong to one of three species.

The data is defined. A basis set (ideally orienting to an interesting
feature) must be provided as an initial orientation. One way to identify
a projection containing interesting features is to apply a guided tour.
In a guided tour, the projection sequence is selected by optimizing an
index function via hill-climbing on the projection space. In this case,
the holes index is selected and applied to standardized flea data. The
holes index is maximized when the projected observations are furthest
from the center. Figure \ref{fig:step0} shows a locally optimized
projection of the data. The left panel displays the reference axes of
the projection basis, a visual indication of the magnitude and direction
each variable contributed to the projection. The right panel shows the
data as projected through the basis set described by the reference axes
(left). Data points are colored and given shape according to the species
(while the guided tour was unsupervised with this information).

@@

\caption[Basis reference axes (left) and projected data (right) of standardized flea data]{Basis reference axes (left) and projected data (right) of standardized flea data. Data points color and shape are mapped to beetle species. The basis was identified by a holes-index guided tour. The variables aede2 and tars1 contribute mostly orthogonal to the other variables.}\label{fig:step0}
\end{figure}
\end{Schunk}

@@

@@

\caption[A radial manual tour of PC4 within the jet cluster]{A radial manual tour of PC4 within the jet cluster. Colored by experiment type: ATLAS7new in green and ATLAS7old in orange. Data from ATLAS7new (green) spans mostly the same space as ALTLAS7old (orange) with no evident difference in cluster structure across varying contributions of PC3. A dynamic version can be viewed at https://nspyrison.netlify.com/thesis/jetcluster\_manualtour\_pc3/.}\label{fig:JetClusterBad}
\end{figure}
\end{Schunk}

@@

A different space is used to explore the DIS cluster; specifically the
first six principal components, which explains 48\% of the variation
contained within the aggregated data (Cook, Laa, and Valencia
\protect\hyperlink{ref-cook_dynamical_2018}{2018}). Radial manual tours
are performed on PC6 and PC2 in figure \ref{fig:DISclusterGood} and
figure \ref{fig:DISclusterBad} respectively.

The selection of the manip variable is important as the manipulation
spaces convey substantially different information. The manual tour of
PC6 offers information about the dimensionality, shape, and orientations
of the different experiment classes. Whereas manipulating the
contributions of PC2 only shows a subset of the dimensionality and shape
information. Manipulating the contributions of PC6 turned out to be much
more insightful than that of PC2. This result might seem
counter-intuitive at first as PC2 should explain much more of the
variation in the data. However, features and structure in the data
regularly reside in finer detail that can be lost when looking only at
static projections.

@@

Tours, which are dynamic linear projection of multivariate data, play an
important role in data visualization; they extend the dimensionality of
visuals while data- and parameter-spaces become ever larger. This
research has modified the algorithm producing manual tours which and has
made this functionality available in package \textbf{spinifex}. The
package adds to \textbf{tourr}, extending the graphics offerings that
can be used to display tours.

Radial manual tours were applied to a dataset across different
experiments of hadronic collisions. The importance of selecting the
correct variable to manipulate is demonstrated by comparing tours of
varying quality. The manual tours convey a better picture of the
structure of the clustering than static linear or non-linear
projections. Giving the full contribution of the manipulation space to
the manip var enables analysts to explore the sensitivity of the
structure with the selected variable. This information can be used by
physicists to identify which experiments are probing which parameter
spaces, which can be indispensable for smarter planning of effort and
funding.

@@

The \textbf{tourr} package provides many other geometric displays with
the \texttt{tourr::display\_*()} family. These geometric options could
be integrated into the \textbf{ggplot2} framework for display on
\textbf{plotly} and \textbf{gganimate}. Additionally, the
\CRANpkg{animation} package Xie et al.
(\protect\hyperlink{ref-xie_animation:_2018}{2018}) could be implemented
for another graphics framework. However, \textbf{animation} builds from
\textbf{base} graphics while \textbf{spinifex} utilizes \textbf{ggplot2}
graphics, a significant paradigm shift.

@@

@@

\leavevmode\hypertarget{ref-kirkpatrick_optimization_1983}{}%
Kirkpatrick, Scott, C. Daniel Gelatt, and Mario P. Vecchi. 1983.
``Optimization by Simulated Annealing.'' \emph{Science} 220 (4598):
671--80.
